Semantic Preserving Bijective Mappings for Representations of Special Functions between Computer Algebra Systems and Document Preparation Systems

### 2018-11-27 main.tex
Avoid one-sentence paragraphs (close #24)

Nonetheless, adopting different notations, in some situations, could also solve potential ambiguities. Consider the two potentially ambiguous examples in~\Cref{tab:amb_ex}. While a scientist would probably just ask for the right interpretation of the first example, \Maple{} automatically computes the first interpretation. On the other hand, \LaTeX{} automatically disambiguates the first example by only recognizing the very next element (single symbols or sequences in curly brackets) for the superscript and therefore displays the second interpretation. The second example should not be misinterpreted, since this notation is the standard interpretation in science for the double factorial. We wrote the second interpretation with parentheses point it out more precisely. However, surprisingly, \Maple{} computes the first interpretation (the factorial of the factorial of $n$) again rather than the common standard interpretation.
\begin{table}[ht]
\centering
\begin{tabular}{lccc}
	\hline
	& Text Format Expression & First Interpretation & Second Interpretation\\
	\hline
	1: & \rule{0pt}{0.9\normalbaselineskip} $4\ \hat{\ }\ 2!$ & $4^{2!}$ & $4^2!$ \\
	2: & $n!!$ & $(n!)!$ & $(n)!!$\\
	\hline
\end{tabular}
\caption{Potentially ambiguous examples of the factorial and double factorial function. One expression in a text format can be interpreted in different ways.}
\label{tab:amb_ex}
\end{table}

@@

	\State \textit{pattern} = \textit{info}.getTranslationPattern();
	\State \textit{translatedMacro} = \textit{pattern}.fillPlaceHolders(\textit{argList});\label{line:fill_pattern}
	\If{\textit{power} is not \NULL}\label{line:shifted_exp}
		\State \textit{translatedMacro}.add(\textit{power});
	\EndIf
	\State \Return\ \textit{translatedMacro};
	\EndProcedure
	\end{algorithmic}
\end{algorithm}

@@

\section{\Maple{} to Semantic \LaTeX{} Translator}\label{sec:backward-translation}
In this section, we will discuss several techniques to get access to the parse tree of \Maple's input. The translation process from this parse tree then follows the same principle as for the forward translations.

Instead of writing a custom \Maple{} syntax parser, we use \Maple's internal data structure to get a syntax tree of the input\footnote{Therefore, a license of \Maple{} is mandatory to perform backward translations. The translator is using the version \Maple{} 2016.}. \Maple{} allows several different input styles. The \texttt{1D} input is mainly used for programming purposes and therefore also used to perform our translations. Internally, \Maple{} uses a \gls*{dag} for syntax trees.

@@

One can access the internal data structure of expressions via the \texttt{ToInert} command, which returns the \inertF. The \inertF{} format is a nested list\footnote{The nested list is a tree representation of a \gls*{dag} that splits nodes with multiple parents into multiple nodes so that each node has only one parent node.} of the internal \gls*{dag} for the given expression. Some of the important types for the nodes are specified in \Cref{tab:maple-types}.
\begin{table}[ht]
\centering
\begin{tabular}{cp{10cm}}
	\hline
	Type & Explanation\\
	\hline
	SUM & Sums. Internally stored with factors for each summand, i.e., `$x+y$' would be stored as `$x \cdot 1 + y \cdot 1$'.\\
	PROD & Products.\\
	EXPSEQ & Expression sequence is a kind of list. The arguments of functions are stored in such sequences.\\
	INTPOS & Positive integers.\\
	INTNEG & Negative integers.\\
	COMPLEX & Complex numbers with real and imaginary part.\\
	FLOAT & Float numbers are stored in the scientific notation with integer values for the exponent $n$ and the significand $m$ in $m \cdot 10^n$.\\
	RATIONAL & Rational numbers are fractions stored in integer values for the numerator and positive integers for the denominator.\\
	POWER & Exponentiation with expressions as base and exponent.\\
	FUNCTION & Function invocation with the name, arguments and attributes of the function.\\
	\hline
\end{tabular}
\caption{A subset of important internal \Maple{} data types. See~\parencite{MAPLE:ProgrammingGuide} for a complete list.}
\label{tab:maple-types}
\end{table}
The translator uses the \texttt{OpenMaple}~\parencite[\S 14.3]{MAPLE:ProgrammingGuide} \gls*{api} for interacting with \Maple's kernel implementation.

@@

\subsection{Round Trip Tests}\label{sec:round-trip}
A round trip test always starts with a valid expression either in semantic \LaTeX{} or in \Maple. A translation from one system to another is called a \textbf{step}. A complete round trip translation (two steps) is called \textbf{one cycle}. A \textbf{fixed point representation} (or short fixed point) in a round trip translation process is a string representation that is identical to all string representations in the following cycles.

Table~\ref{tab:fixpoint} illustrates an example of a round trip test which reaches a fixed point. The test formula is
\begin{equation}
\frac{\cos\left(a\Theta\right)}{2}.
\end{equation}

@@

There is currently only one exception known where a round trip test does not reach a fixed point representation: Legendre's incomplete elliptic integrals~\parencite[(19.2.4-7)]{NIST:DLMF} are defined with the amplitude $\phi$ in the first argument in the \DLMF, while \Maple{} takes the trigonometric sine of the amplitude as the first argument. Therefore, the forward and backward translations are defined as
\begin{eqnarray}
\verb|\EllIntF@{\phi}{k}| & \overset{\langMaple}{\mapsto} & \verb|EllipticF(sin(phi),k)|,\\
\verb|\EllIntF@{\asin@{\phi}}{k}| & \overset{\langMaple}{\mapsfrom} & \verb|EllipticF(phi,k)|.
\end{eqnarray}
The round-trip translations produce infinite chains of sine and inverse sine calls because there are no evaluations involved.

@@

A theoretical improvement for the numerical tests would be to perform the tests for specific, considered as critical, values with respect to the involved functions. J.~C.~Beamont et al.~tested  identities for multivalued elementary functions by choosing sample points from a region with respect to the branch cuts of the functions~\parencite{DBLP:journals/aaecc/BeaumontBDP07}. We assume that choosing sample points from those regions, significantly improves the results of our numerical tests.

Increasing the amount of translatable formulae in the \gls*{dlmf} and improving the verificiation techniques are parts of ongoing projects.

>>>>

Nonetheless, adopting different notations, in some situations, could also solve potential ambiguities. Consider the two potentially ambiguous examples in~\Cref{tab:amb_ex}. While a scientist would probably just ask for the right interpretation of the first example, \Maple{} automatically computes the first interpretation. On the other hand, \LaTeX{} automatically disambiguates the first example by only recognizing the very next element (single symbols or sequences in curly brackets) for the superscript and therefore displays the second interpretation. The second example should not be misinterpreted, since this notation is the standard interpretation in science for the double factorial. We wrote the second interpretation with parentheses point it out more precisely. However, surprisingly, \Maple{} computes the first interpretation (the factorial of the factorial of $n$) again rather than the common standard interpretation.
\begin{table}[ht]
\centering
\begin{tabular}{lccc}
	\hline
	& Text Format Expression & First Interpretation & Second Interpretation\\
	\hline
	1: & \rule{0pt}{0.9\normalbaselineskip} $4\ \hat{\ }\ 2!$ & $4^{2!}$ & $4^2!$ \\
	2: & $n!!$ & $(n!)!$ & $(n)!!$\\
	\hline
\end{tabular}
\caption{Potentially ambiguous examples of the factorial and double factorial function. One expression in a text format can be interpreted in different ways.}
\label{tab:amb_ex}
\end{table}
\vspace*{-0.5cm}

@@

	\State \textit{pattern} = \textit{info}.getTranslationPattern();
	\State \textit{translatedMacro} = \textit{pattern}.fillPlaceHolders(\textit{argList});\label{line:fill_pattern}
	\If{\textit{power} is not \NULL}\label{line:shifted_exp}
		\State \textit{translatedMacro}.add(\textit{power});
	\EndIf
	\State \Return\ \textit{translatedMacro};
	\EndProcedure
	\end{algorithmic}
\end{algorithm}
\vspace*{-0.2cm}

@@

\section{\Maple{} to Semantic \LaTeX{} Translator}\label{sec:backward-translation}
In this section, we will discuss several techniques to get access to the parse tree of \Maple's input. The translation process from this parse tree then follows the same principle as for the forward translations. Instead of writing a custom \Maple{} syntax parser, we use \Maple's internal data structure to get a syntax tree of the input\footnote{Therefore, a license of \Maple{} is mandatory to perform backward translations. The translator is using the version \Maple{} 2016.}. \Maple{} allows several different input styles. The \texttt{1D} input is mainly used for programming purposes and therefore also used to perform our translations. Internally, \Maple{} uses a \gls*{dag} for syntax trees.

@@

One can access the internal data structure of expressions via the \texttt{ToInert} command, which returns the \inertF. The \inertF{} format is a nested list\footnote{The nested list is a tree representation of a \gls*{dag} that splits nodes with multiple parents into multiple nodes so that each node has only one parent node.} of the internal \gls*{dag} for the given expression. Some of the important types for the nodes are specified in \Cref{tab:maple-types}. The translator uses the \texttt{OpenMaple}~\parencite[\S 14.3]{MAPLE:ProgrammingGuide} \gls*{api} for interacting with \Maple's kernel implementation.

\begin{table}[ht]
\centering
\begin{tabular}{cp{10cm}}
	\hline
	Type & Explanation\\
	\hline
	SUM & Sums. Internally stored with factors for each summand, i.e., `$x+y$' would be stored as `$x \cdot 1 + y \cdot 1$'.\\
	PROD & Products.\\
	EXPSEQ & Expression sequence is a kind of list. The arguments of functions are stored in such sequences.\\
	INTPOS & Positive integers.\\
	INTNEG & Negative integers.\\
	COMPLEX & Complex numbers with real and imaginary part.\\
	FLOAT & Float numbers are stored in the scientific notation with integer values for the exponent $n$ and the significand $m$ in $m \cdot 10^n$.\\
	RATIONAL & Rational numbers are fractions stored in integer values for the numerator and positive integers for the denominator.\\
	POWER & Exponentiation with expressions as base and exponent.\\
	FUNCTION & Function invocation with the name, arguments and attributes of the function.\\
	\hline
\end{tabular}
\caption{A subset of important internal \Maple{} data types. See~\parencite{MAPLE:ProgrammingGuide} for a complete list.}
\label{tab:maple-types}
\end{table}

@@

\subsection{Round Trip Tests}\label{sec:round-trip}
A round trip test always starts with a valid expression either in semantic \LaTeX{} or in \Maple. A translation from one system to another is called a \textbf{step}. A complete round trip translation (two steps) is called \textbf{one cycle}. A \textbf{fixed point representation} (or short fixed point) in a round trip translation process is a string representation that is identical to all string representations in the following cycles. Table~\ref{tab:fixpoint} illustrates an example of a round trip test which reaches a fixed point for the formula
\begin{equation}
\frac{\cos\left(a\Theta\right)}{2}.
\end{equation}

@@

There is currently only one exception known where a round trip test does not reach a fixed point representation: Legendre's incomplete elliptic integrals~\parencite[(19.2.4-7)]{NIST:DLMF} are defined with the amplitude $\phi$ in the first argument in the \DLMF, while \Maple{} takes the trigonometric sine of the amplitude as the first argument. Therefore, the forward and backward translations are defined as
\begin{eqnarray}
\verb|\EllIntF@{\phi}{k}| & \overset{\langMaple}{\mapsto} & \verb|EllipticF(sin(phi),k)|,\\
\verb|\EllIntF@{\asin@{\phi}}{k}| & \overset{\langMaple}{\mapsfrom} & \verb|EllipticF(phi,k)|,
\end{eqnarray}
and the round-trip translations produce infinite chains of sine and inverse sine calls because there are no evaluations involved.

@@

A theoretical improvement for the numerical tests would be to perform the tests for specific, considered as critical, values with respect to the involved functions. J.~C.~Beamont et al.~tested  identities for multivalued elementary functions by choosing sample points from a region with respect to the branch cuts of the functions~\parencite{DBLP:journals/aaecc/BeaumontBDP07}. We assume that choosing sample points from those regions, significantly improves the results of our numerical tests. Increasing the amount of translatable formulae in the \gls*{dlmf} and improving the verificiation techniques are parts of ongoing projects.
